
\begin{thebibliography}{99}

\bibitem{jurafsky}
Jurafsky D., Martin J. H.
\textit{Speech and Language Processing}.
3rd ed. --- Draft. Stanford University, 2023.

\bibitem{vaswani}
Vaswani A., Shazeer N., Parmar N. et al.
Attention Is All You Need //
\textit{Advances in Neural Information Processing Systems (NeurIPS)}. --- 2017.

\bibitem{asr_survey}
Hinton G., Deng L., Yu D. et al.
Deep Neural Networks for Acoustic Modeling in Speech Recognition //
\textit{IEEE Signal Processing Magazine}. --- 2012.

\bibitem{goodfellow_dl}
Goodfellow I., Bengio Y., Courville A.
\textit{Deep Learning}.
MIT Press, 2016.

\bibitem{real_time}
Ma M., Huang L., Xiong H. et al.
STACL: Simultaneous Translation with Implicit Anticipation and Controllable
Latency //
\textit{Proceedings of the Annual Meeting of the Association for Computational
Linguistics (ACL)}. --- 2019.

\bibitem{papineni_bleu}
Papineni K., Roukos S., Ward T., Zhu W.-J.
BLEU: a Method for Automatic Evaluation of Machine Translation //
\textit{Proceedings of the 40th Annual Meeting of the Association for Computational
Linguistics (ACL)}. --- 2002.

\bibitem{bert_score}
Zhang T., Kishore V., Wu F., Weinberger K. Q., Artzi Y.
BERTScore: Evaluating Text Generation with BERT //
\textit{Proceedings of the International Conference on Learning Representations
(ICLR)}. --- 2020.

\bibitem{banerjee_meteor}
Banerjee S., Lavie A.
METEOR: An Automatic Metric for MT Evaluation with Improved Correlation with
Human Judgments //
\textit{Proceedings of the ACL Workshop on Intrinsic and Extrinsic Evaluation
Measures for Machine Translation and/or Summarization}. --- 2005.

\bibitem{whisper}
Radford A., Kim J. W., Xu T. et al.
Robust Speech Recognition via Large-Scale Weak Supervision //
\textit{Proceedings of the International Conference on Machine Learning
(ICML)}. --- 2023.

\bibitem{post_bleu}
Post M.
A Call for Clarity in Reporting BLEU Scores //
\textit{Proceedings of the Third Conference on Machine Translation (WMT)}. --- 2018.

\bibitem{opus_mt}
Tiedemann J., Thottingal S.
OPUS-MT --- Building open translation services for the World //
\textit{Proceedings of the 22nd Annual Conference of the European Association
for Machine Translation (EAMT)}. --- 2020.

\bibitem{nllb}
NLLB Team, Costa-jussà M. R. et al.
No Language Left Behind: Scaling Human-Centered Machine Translation //
\textit{arXiv preprint arXiv:2207.04672}. --- 2022.

\bibitem{seamlessm4t}
Barrault L., Chung Y.-A., Meglioli M. C. et al.
SeamlessM4T --- Massively Multilingual \& Multimodal Machine Translation //
\textit{arXiv preprint arXiv:2308.11596}. --- 2023.

\bibitem{gpt4}
OpenAI.
GPT-4 Technical Report //
\textit{arXiv preprint arXiv:2303.08774}. --- 2023.

\bibitem{mann_whitney}
Mann H. B., Whitney D. R.
On a Test of Whether One of Two Random Variables is Stochastically Larger
than the Other //
\textit{The Annals of Mathematical Statistics}. --- 1947. --- Vol.~18, No.~1.
--- P.~50--60.

\bibitem{spearman}
Spearman C.
The Proof and Measurement of Association between Two Things //
\textit{The American Journal of Psychology}. --- 1904. --- Vol.~15, No.~1.
--- P.~72--101.

\bibitem{marian_nmt}
Junczys-Dowmunt M., Grundkiewicz R., Dwojak T. et al.
Marian: Fast Neural Machine Translation in C++ //
\textit{Proceedings of ACL 2018, System Demonstrations}. --- 2018.

\bibitem{ctranslate2}
Klein G., Hernandez F., Nguyen V.
CTranslate2: Efficient Inference with Transformer Models //
\textit{GitHub repository}. --- 2020.

\bibitem{flores}
Goyal N., Gao C., Chaudhary V. et al.
The FLORES-101 Evaluation Benchmark for Low-Resource and Multilingual
Machine Translation //
\textit{Transactions of the Association for Computational Linguistics}. --- 2022.

\bibitem{cascade_st}
Sperber M., Paulik M.
Speech Translation and the End-to-End Promise: Taking Stock of Where
We Are //
\textit{Proceedings of the 58th Annual Meeting of the Association for
Computational Linguistics (ACL)}. --- 2020.

\bibitem{iranzo_cascade}
Iranzo-Sánchez J., Silvestre-Cerdà J. A., Jorge J. et al.
Direct Speech Translation for Automatic Subtitling //
\textit{Proceedings of Interspeech}. --- 2021.

\bibitem{bentivogli_cascade}
Bentivogli L., Cettolo M., Federico M., Cattoni R.
Machine Translation Quality: Human vs. Cascade Effects in End-to-End
Speech Translation //
\textit{Proceedings of the International Workshop on Spoken Language
Translation (IWSLT)}. --- 2021.

\bibitem{mqm}
Lommel A., Uszkoreit H., Burchardt A.
Multidimensional Quality Metrics (MQM): A Framework for Declaring and
Describing Translation Quality Metrics //
\textit{Tradumàtica: tecnologies de la traducció}. --- 2014. --- No.~12.
--- P.~455--459.

\bibitem{llm_mt}
Zhu W., Liu H., Dong Q. et al.
Multilingual Machine Translation with Large Language Models: Empirical
Results and Analysis //
\textit{Findings of NAACL}. --- 2024.

\bibitem{wav2vec}
Baevski A., Zhou Y., Mohamed A., Auli M.
wav2vec 2.0: A Framework for Self-Supervised Learning of Speech
Representations //
\textit{Advances in Neural Information Processing Systems (NeurIPS)}. --- 2020.

\bibitem{opus_corpus}
Tiedemann J.
Parallel Data, Tools and Interfaces in OPUS //
\textit{Proceedings of the 8th International Conference on Language
Resources and Evaluation (LREC)}. --- 2012.

\bibitem{wer_metric}
Morris A., Maier V., Green P.
From WER and RIL to MER and WIL: Improved Evaluation Measures for
Connected Speech Recognition //
\textit{Proceedings of Interspeech}. --- 2004.

\bibitem{low_resource_mt}
Zoph B., Yuret D., May J., Knight K.
Transfer Learning for Low-Resource Neural Machine Translation //
\textit{Proceedings of the Conference on Empirical Methods in Natural
Language Processing (EMNLP)}. --- 2016.

\bibitem{hendy_llm_mt}
Hendy A., Abdelrehim M., Sharaf A. et al.
How Good Are GPT Models at Machine Translation? A Comprehensive Evaluation //
\textit{arXiv preprint arXiv:2302.09210}. --- 2023.

\bibitem{domain_adaptation}
Chu C., Wang R.
A Survey of Domain Adaptation for Machine Translation //
\textit{Journal of Information Processing}. --- 2020. --- Vol.~28.
--- P.~413--426.

\bibitem{uk_nlp}
Romanyuk A., Ignatova T.
Ukrainian Language Resources and Tools: The State of the Art //
\textit{Proceedings of the Workshop on Language Technology for
Less-Resourced Languages (LT4ALL)}. --- 2019.

\bibitem{error_propagation}
Ruiz N., Federico M.
Assessing the Impact of Speech Recognition Errors on Machine Translation
Quality //
\textit{Proceedings of the 11th Conference of the Association for Machine
Translation in the Americas (AMTA)}. --- 2014.

\bibitem{st_survey}
Anastasopoulos A., Chiang D.
A Survey on End-to-End Speech Translation //
\textit{arXiv preprint arXiv:2104.06494}. --- 2022.

\end{thebibliography}